---
title: "Preprocessing an HT-SELEX study"
author: "Marco Rotanegroni"
date: "`r Sys.Date()`"
output: BiocStyle::html_document
vignette: >
  %\\VignetteIndexEntry{Preprocessing an HT-SELEX study}
  %\\VignetteEngine{knitr::rmarkdown}
  %\\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Introduction

High-throughput SELEX experiments enrich binding sequences over successive
rounds, but public deposits commonly expose raw reads without a uniform record
of primer layout, trimming decisions, or the relationship between rounds.
`selexprepR` provides a reproducible preprocessing layer for this setting. It
uses `Biostrings` sequence containers, returns a sparse
`SummarizedExperiment`, and stores its inference, extraction, quality-control,
and provenance records alongside the assay.

The package complements general FASTQ infrastructure in Bioconductor. Its
domain-specific contribution is conservative primer inference and a manifest
that makes the preprocessing decision inspectable rather than implicit.
The existing `SELEX` package provides a complementary analysis workflow based
on its own sample annotation model. `selexprepR` focuses on raw public-read
acquisition, primer inference, and a `SummarizedExperiment` output that can be
used by downstream Bioconductor tooling.

# Installation

```{r installation, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}
BiocManager::install("selexprepR")
```

# A complete in-memory workflow

For a self-contained example, create three synthetic rounds with fixed 5-prime
and 3-prime constant regions. The variable regions below are distinct so the
count matrix has a realistic multi-sequence shape.

```{r synthetic-data}
library(selexprepR)

primer_5p <- "GGTAATACGACTCACTATAGGG"
primer_3p <- "CCATGCATGCATGCATGCAT"
bases <- c("A", "C", "G", "T")

make_round <- function(round, n = 500L, width = 20L) {
    variable_regions <- vapply(seq.int(0L, n - 1L), function(index) {
        paste0(bases[(index %/% 4L^(0L:(width - 1L))) %% 4L + 1L], collapse = "")
    }, character(1))
    paste0(primer_5p, variable_regions, primer_3p)
}

reads_by_round <- stats::setNames(lapply(0:2, make_round), sprintf("round_%02d", 0:2))
experiment <- run_selexprep(reads_by_round, low_total_reads = 0)
experiment
```

The counts assay has one column per SELEX round and one row per recovered
variable sequence. It is sparse, so conventional SELEX datasets with many
unobserved sequence-round combinations remain compact.

```{r inspect-counts}
counts <- SummarizedExperiment::assay(experiment, "counts")
Matrix::colSums(counts)
head(SummarizedExperiment::rowData(experiment))
```

The library report records the inferred constant regions and the decision about
whether full inserts were recovered. The QC object summarizes every round and
lists explicit flags instead of silently discarding a concern.

```{r inspect-provenance}
metadata <- S4Vectors::metadata(experiment)
metadata$library_report[c("primer_5p", "primer_3p", "extraction_mode", "confidence")]
metadata$qc$per_round
metadata$qc$flags
metadata$manifest$manifest_version
```

# Reading FASTQ and public metadata

For local FASTQ files, the file bridge assembles rounds and retains file sizes,
MD5 values, and SHA-256 hashes in the experiment metadata and manifest.

```{r fastq-input, eval=FALSE}
experiment <- run_selexprep_files(c(
    round_00 = "round_00.fastq.gz",
    round_01 = "round_01.fastq.gz"
))
```

When several rounds share one inline-barcoded R1 stream, provide the barcode
map explicitly. Barcodes are checked for sufficient pairwise Hamming distance;
the package does not infer them.

```{r demultiplex, eval=FALSE}
inputs <- selexprep_demultiplex(
    read_selexprep_fastq("multiplexed.fastq.gz"),
    c(AAAAA = 0L, TTTTT = 1L)
)
experiment <- run_selexprep(inputs)
```

The catalog is distributed as a package dataset and can be queried without a
network connection. Its metadata records the exact snapshot and immutable
source locations.

```{r public-catalog}
data("selexprep_public_catalog", package = "selexprepR")
selexprep_catalog("FGF-9")
S4Vectors::metadata(selexprep_public_catalog)$snapshot_version
```

ENA inspection and download are separate, so the planned files and checksums
can be reviewed before any data transfer.

```{r ena-public-data, eval=FALSE}
inspection <- inspect_selexprep_accession("PRJDB19098")
fetch_selexprep_reads(inspection, "raw/PRJDB19098")
```

# References

Tuerk C, Gold L (1990). Systematic evolution of ligands by exponential
enrichment: RNA ligands to bacteriophage T4 DNA polymerase. *Science*,
249(4968), 505-510. doi:10.1126/science.2200121.

# Session information

```{r session-info}
sessionInfo()
```
